---
title: "Data Clean: Peer Assessment 2 - ReadMe"
output: 
  html_document:
    keep_md: true
---

## Overview
The run_analysis.R script file is used to generate a tiday data set with the average of each variable for each activity and each subject.  
The data sets are created by merging training and the test sets generated by UCI's Human Activity Recognition Using Smartphones project.


## data sets built from data files
=========================================

- Activity Label 
- Activity
- Features
- Measures
- Subject

## tidy data output
- subject_id
- activity
- average by subject and activity 



#### libraries used
```{r message=FALSE}

library(reshape2)
library(tidyr)
library(plyr)
library(dplyr)

```


#### Read data file to create a factor variable in activity data frame
      
```{r}
activity_labels <- read.table("./Data/UCI HAR Dataset/activity_labels.txt",
                              header = FALSE,
                              stringsAsFactors=FALSE,
                              col.names=c("activity_id","label"))
                              
```


#### Read two subject data files and  merge them into one data frame with 10299 obs, 1 var
* add unique row index as row identifier
```{r}
s1 <- read.table("./Data/UCI HAR Dataset/test/subject_test.txt",
                 header = FALSE,
                 stringsAsFactors=FALSE,
                 col.names=c("subject_id"))
s2 <- read.table("./Data/UCI HAR Dataset/train/subject_train.txt",
                 header = FALSE,
                 stringsAsFactors=FALSE,
                 col.names=c("subject_id"))
subject <- rbind(s1, s2)  
subject$id <- seq(along=subject$subject_id)  

```


#### read two data files and combine into one data frame
* combined data frame has 10299 obs, 2 vars - activity_id and index_id
```{r}
y1 <- read.table("./Data/UCI HAR Dataset/test/y_test.txt",
                 header = FALSE,
                 stringsAsFactors=FALSE,
                 col.names=c("activity_id"))
y2 <- read.table("./Data/UCI HAR Dataset/train/y_train.txt",
                 header = FALSE,
                 stringsAsFactors=FALSE,
                 col.names=c("activity_id"))
activity <- rbind(y1, y2)
activity$id <- seq(along=activity$activity_id)
```


#### change activity_id to a factor variable with level from activity_labelss dataset
```{r}
f <-  as.factor(activity$activity_id)
levels(f) <- activity_labels$label
activity$activity_id <- f
```



#### build  feature dataset which used to filter mean and std computes
```{r}
features <- read.table("./Data/UCI HAR Dataset/features.txt",
                       header = FALSE,
                       stringsAsFactors=FALSE,
                       col.names=c("feature_id", "descr"))
```                    
                       
                       
#### split a column with three different variable into 3 different columns for each variable
* features now has 66 obs and 2 vars - feature_id, computemethod - mean, std

```{r}
f_temp <- colsplit(string=features[,2],
                     pattern="-",
                     names=c("feature", "computemethod", "xyz"))
features$computemethod <- sub("()", "", f_temp$computemethod, fixed=TRUE)

features <- features %>% filter(computemethod  %in% c("mean", "std"))
```


##### Build one measure data set from two data files.  combined set has 10299 obs, 56 variables
```{r}
x1 <- read.table("./Data/UCI HAR Dataset/test/x_test.txt", header = FALSE , stringsAsFactors=FALSE)
x2 <- read.table("./Data/UCI HAR Dataset/train/x_train.txt", header = FALSE, stringsAsFactors=FALSE)
measure <- rbind(x1, x2)  
```


#### rebuild variable names from existing genetic column names V1, V2, ...
* the numeric colnum names without V is mapped to feature_id
```{r}
colnames(measure) <- gsub("^V", "", colnames(measure))
```


add row identifier index using seq() function.  
do not use seq's along parameter.  it gives wrong result when ref values are < 0 
10299 obs, 66 variables + 1 = 67 vars including id field for join with other datasets
```{r}
measure <- measure[,features$feature_id]
measure$id <- seq(1,nrow(measure), by=1)  
```


#### Final tidy data shaping
* join subject, activity, measure datasets by id
* tidy data set with 10299 obs, 69 vars including id, subject_id, activity_id, and measure var(1-66)
```{r}

dflist <- list(as.data.frame(subject), as.data.frame(activity), as.data.frame(measure))
tidydata <- join_all(dflist)

# change subject_id to factor variable
tidydata$subject_id <- as.factor(tidydata$subject_id)

## include only needed columns to compute avg for each activity
tidydata$id <- NULL
tidydata$avg <- rowMeans(tidydata[,3:68])

## only keep the relevant info to create cross tabs and convert it back to data frame for further shaping
tidydata <- tidydata[,c("subject_id", "activity_id", "avg")]

## build cross tab for avg grouped by subject_id, activity_id
xt <- xtabs(avg ~ subject_id + activity_id, data=tidydata)

## change back to data frame to handle group by and sort for the output file
xt <- as.data.frame(xt)

## group and sort the output data by subject and activity
output <- xt %>% group_by(subject_id, activity_id) %>% arrange(subject_id, activity_id)
```


#### Write out the result to a file
make sure to include col.names and row.names paraeters to FALSE. Column names incorrect when set to TRUE
```{r}
write.table(output, file = "./Data/tidydata_output.txt", col.names=FALSE, row.names=FALSE)

```








